### 2021-03-01 manuscript/v5_submission_revision/manuscript_tex4submission.tex
add new folders for revision, test latexdiff

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{medRxiv} % Sets Journal name

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc header
\usepackage{geometry}
% \usepackage[numbers,sort&compress]{natbib}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\begin{frontmatter}

  \title{Causal analysis of COVID-19 observational data in German districts
reveals effects of mobility, awareness, and temperature}
    \author[Zi]{Edgar Steiger}
   \ead{esteiger@zi.de} 
    \author[Zi]{Tobias Mußgnug}
   \ead{tmussgnug@zi.de} 
    \author[Zi]{Lars Eric Kroll}
   \ead{lkroll@zi.de} 
      \address[Zi]{Central Research Institute of Ambulatory Health Care in Germany (Zi),
Salzufer 8, D-10587 Berlin, Germany}

  \begin{abstract}
  Mobility, awareness, and weather are suspected to be causal drivers for
  new cases of COVID-19 infection. Correcting for possible confounders, we
  estimated their causal effects on reported case numbers. To this end, we
  used a directed acyclic graph (DAG) as a graphical representation of the
  hypothesized causal effects of the aforementioned determinants on new
  reported cases of COVID-19. Based on this, we computed valid adjustment
  sets of the possible confounding factors. We collected data for Germany
  from publicly available sources (e.g.~Robert Koch Institute, Germany's
  National Meteorological Service, Google) for 401 German districts over
  the period of 15 February to 8 July 2020, and estimated total causal
  effects based on our DAG analysis by negative binomial regression. Our
  analysis revealed favorable causal effects of increasing temperature,
  increased public mobility for essential shopping (grocery and pharmacy),
  and awareness measured by COVID-19 burden, all of them reducing the
  outcome of newly reported COVID-19 cases. Conversely, we saw adverse
  effects of public mobility in retail and recreational areas, awareness
  measured by searches for ``corona'' in Google, and higher rainfall,
  leading to an increase in new COVID-19 cases. This comprehensive causal
  analysis of a variety of determinants affecting COVID-19 progression
  gives strong evidence for the driving forces of mobility, public
  awareness, and temperature, whose implications need to be taken into
  account for future decisions regarding pandemic management.
  \end{abstract}

 \end{frontmatter}

\section{Introduction}\label{introduction}

As the COVID-19 pandemic progresses, research on mechanisms behind the
transmission of SARS-CoV-2 shows conflicting evidence
\citep{who2020report, Chinazzi395, guan2020clinical}. While effects of
mobility have been extensively discussed, less is known on other factors
such as changing awareness in the population
\citep{higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020}
or the effects of temperature
\citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020}.
A limiting factor in many studies is the lack of a causal approach to
assess the causal contributions of various factors
\citep{Greenland1999}. This can lead to distorted estimates of the
causal factors with observational data
\citep{Greenland1999, schipf_directed_2011, textor_robust_2017}.

@@

Another problem with assessing the effects of various determinants of
SARS-CoV-2 spread is the heterogeneity of the countries and regions
examined for example in the Johns Hopkins University (JHU) COVID-19
database \citep{jhucovid19db2020}. The comparison of time series of case
numbers from different countries and observational periods can be
strongly distorted by different factors like testing capacities and
regional variations.

Our objective is to provide estimates of the causal effects of the main
drivers of the pandemic with reduced bias. We conducted a scoping review
of the available studies regarding signaling pathways and determinants
of the spread of SARS-CoV-2 infections and the reported new COVID-19
cases. Then we integrated the current findings into a directed acyclic
graph for the progress of the pandemic at the regional level. Using the
resulting model and the do-calculus we found identifiable effects
without blocked causal paths whose effects can be analyzed with
observational data \citep{Pearl_2014}. We used regional time series data
of all German districts (\(401\)) from various publicly available
sources to analyze these questions on a regional level. Germany is a
good choice in this regard, because it has ample data on contributing
factors on the regional level and has had high testing and treatment
capacities from early on in the pandemic.

\section{Causal Model}\label{causal-model}

We used a directed acyclic graph (DAG)
\citep{schipf_directed_2011, textor_robust_2017} as a tool to analyze
the causal relationships between several exposures and SARS-CoV-2
spread. To get an overview on published associations, a scoping review
was conducted from 20th to 22nd of May 2020 within Pubmed and Google
scholar. Restrictions were applied to English and German language and
the publication date in the last one year. The following search terms
were applied to abstracts and title in Pubmed (``COVID-19'' OR
``COVID19'' OR ``Corona'' OR ``Coronavirus'' OR ``SARS-CoV-2'') and
connected separately in each case with the exposure variables
(``mobility'', ``public awareness'', ``awareness'', ``google
trends'',``ambient temperature'', ``temperature''). For ``mobility'', we
analyzed \(n=8\) studies, \(N=103\) were scanned in Pubmed, together
with the first ten pages (100 results) in Google scholar
(``awareness''/``public awareness''/``google trends'' \(n=9\),
\(N=215\); ``temperature''/``ambient temperature'' \(n=16\), \(N=235\)).
We integrated these findings where possible into the construction of our
DAG, which can be seen in Figure \ref{fig:dag-covid-19}.

\begin{figure}
\centering
\includegraphics{figures/f_full_covid19_dag.eps}
\caption{\label{fig:dag-covid-19}DAG of determinants of reported COVID-19
cases on the district level. Unobserved variables are light gray,
variables marked with an asterisk (*) are confounded by
weekday/holiday.}
\end{figure}

A number of studies report a strong association of \textbf{mobility}
restrictions on the number of new COVID-19 cases: Restrictive measures
(e.g. ``stay-at-home'' orders, travel bans, or school closures) are
shown to possibly reduce the COVID-19 incidence
\citep{chang_modeling_2020, Chinazzi395, fowler_effect_2020, kraemer_effect_2020, lasry_timing_2020, linka_outbreak_2020, mazzoli_effects_2020, xiong_data-driven_2020}.
However, some studies point out the combination of various
non-pharmaceutical interventions (NPIs) is decisive to prevent new
infections \citep{juni_impact_2020, lai_effect_2020}.

Google Trends \citep{google_trends} data can be used as a tool to get
insights into public interest (\textbf{awareness}) in the coronavirus
disease. Several recent studies imply a connection of relative search
volumes (RSV) indices and reported new COVID-19 cases
\citep{ayyoubzadeh_predicting_2020, effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, lin_google_2020, mavragani_tracking_2020, walker_use_2020, yuan_trends_2020, zhou_effects_2020}.
Some search terms e.g. ``COVID-19'' or ``coronavirus'' predated newly
infected cases/total number of cases by roughly 7 to 14 days for
different countries
\citep{effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020}.
Additionally, we acknowledged that individual risk-aware behavior might
be a reaction to the current COVID-19 burden (measured as reported cases
at the day of exposure).

Mixed evidence is available regarding the effect of
\textbf{temperature}: On the one hand several papers report an
association between increase in temperature and decrease in newly
infected COVID-19 cases
\citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
On the other hand, also the opposite has been found
\citep{auler_evidence_2020, xie_association_2020}. Some studies found no
association at all
\citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020}.
It should be noted that few studies considered other confounding
variables than meteorological ones (especially age and population
density among others
\citep{briz-redon_spatio-temporal_2020, juni_impact_2020, wu_effects_2020}).
In addition, the transferability of results between different climate
zones is questionable. To avoid possible bias caused by weather
variables other than temperature, we included rain, wind, and humidity
in our model.

When investigating causal determinants of SARS-CoV-2 infections, a
number of confounders have to be considered. Well-known risk factors for
SARS-CoV-2 as well as for other infections are demographic factors such
as age, gender, socio-economic status (SES), population density, and
foreign citizenship/ethnicity
\citep{de2020risk, Dragano2020.06.17.20133918, jhucovid19db2020}. In
Germany along with other countries (i.e.~Brazil, USA, or the UK),
populist parties or politicians and their electorate tend to be more
sceptical about effects of containment measures than the other part of
the electorate \citep{dohle_wingen_schreiber_2020, engle_staying_2020}.
Therefore we considered both ``right-wing populist party votes'' and
``voter turnout'' as possible confounders. Public health interventions
were also taken into account (contact restrictions, school closures
etc.), as their implementation showed strong correlations with
controlling the spread of SARS-CoV-2
\citep{cowling2020impact, juni_impact_2020, lai_effect_2020}. To avoid
bias due to reporting delay of case numbers we had to include weekday
and German holidays. We include some unobserved variables in our DAG
(e.g. ``Herd immunity''), too. Please note that ``Exposure to
SARS-CoV-2'' is itself an unobserved variable: German case numbers are
reported with delay after date of exposure and symptom onset.
\emph{Exposure to the virus} should not be confused with the formal
\emph{exposure variables} of the DAG (mobility, awareness, temperature).

\section{Data}\label{data}

We collected and aggregated data on reported COVID-19 cases, regional
socio-demographic factors, weather, and general mobility on district and
state level in Germany for the period of 15 February 2020 to 8 July
2020. Our observation period for the outcome consisted of all dates from
23 February 2020 to 8 July 2020 (\(T=137\)), since we used a lag of 8
days for all confounders. We did not exclude any states or districts
(\(K=401\)). We analyzed the daily reported number of new cases as
outcome (\(K\cdot T=54{\ }937\) observations). The set of possible
predictors was derived from our causal DAG (see Table
\ref{tab:model-vars} and Figure \ref{fig:dag-covid-19}). Due to
modelling and data limitations, some of the predictors were unobserved
or were modelled as a construct consisting of several variables. For our
causal analysis, we computed adjustment sets in three different
scenarios for separate exposures within the DAG: i) mobility of
population, ii) awareness of COVID-19 (i.e.~Google searches for
``corona''), iii) weather (i.e.~temperature).

\subsection{Variables}\label{variables}

We downloaded German daily case numbers on district level reported by
Robert Koch Institute (RKI, \citep{casenumbers_rki}, acquired on 12 July
2020) and aggregated them by date. The number of daily active cases for
day \(d\) was derived by subtracting the total number of reported cases
on day \(d\) and day \(d-14\) (14 days as a conservative estimate for
the infectious period, which corresponds here to the required quarantine
time in Germany).

To assess the mobility of the German population, we used data publicly
available on German state level from Google \citep{google_mobility}.
Measurements are daily relative changes of mobility in percent compared
to the period of 3 January 2020 to 6 February 2020. Missing values
(\(25\) out of \(13{\ }488\)) were imputed with value \(0\) and the
state level measurements were passed onto districts within the
corresponding state. Google mobility data was available for six
different sectors of daily life (``retail and recreation'', ``grocery
and pharmacy'', ``parks'', ``transit stations'', ``workplaces'',
``residential'') which means that ``mobility'' is a construct consisting
of several variables. All variables but ``residential'' mobility are
relative changes of daily visitor numbers to the corresponding sectors
compared to the reference period. ``Residential'' mobility is the
relative change of daily time spent at residential areas.

The notion of awareness in the population of COVID-19 describes the
general state of alertness about the new infectious disease. As such, it
was hard to measure directly. As a proxy, we used the relative interest
in the topic term ``corona'' as indicated by Google searches. The daily
data was available on state level \citep{google_trends} and passed onto
district level. As a second proxy for awareness, we used the daily
reported number of COVID-19 cases on the day of the exposure: Since
media reported case numbers prominently, we assumed that this could
reflect individual awareness, too.

We constructed daily weather from four variables (``temperature'',
``rainfall'', ``humidity'', ``wind''). Weather data was downloaded from
Deutscher Wetterdienst (DWD, \citep{dwd_weather}) for all weather
stations in Germany below 1000 meters altitude with daily records for
our observation period. District level daily weather data was aggregated
per district by averaging the data from the three nearest weather
stations (which includes weather stations inside the district). Missing
values were imputed with mean values (\(n=59\) for wind).

>>>>

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% my packages and configs
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{url}
\def\UrlBreaks{\do\/\do-}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Causal analysis of COVID-19 observational data in German districts
reveals effects of mobility, awareness, and temperature} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Edgar Steiger\textsuperscript{1*},
Tobias Mußgnug\textsuperscript{1},
Lars Eric Kroll\textsuperscript{1}%,
%Name4 Surname\textsuperscript{2},
%Name5 Surname\textsuperscript{2\ddag},
%Name6 Surname\textsuperscript{2\ddag},
%Name7 Surname\textsuperscript{1,2,3*},
%with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Central Research Institute of Ambulatory Health Care in Germany (Zi),
Salzufer 8, D-10587 Berlin, Germany
\\
%\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
%\\
%\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
%\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
%\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
%\ddag These authors also contributed equally to this work.

% Current address notes
%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
%\dag Deceased

% Group/Consortium Author Note
%\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* esteiger@zi.de

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Mobility, awareness, and weather are suspected to be causal drivers for new cases of COVID-19 infection. Correcting for possible confounders, we estimated their causal effects on reported case numbers. To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the aforementioned determinants on new reported cases of COVID-19. Based on this, we computed valid adjustment sets of the possible confounding factors. We collected data for Germany from publicly available sources (e.g.~Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression. Our analysis revealed favorable causal effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy), and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases. Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for ``corona'' in Google, and higher rainfall, leading to an increase in new COVID-19 cases. This comprehensive causal analysis of a variety of determinants affecting COVID-19 progression gives evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%\section*{Author summary}
%Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}\label{introduction}
As the COVID-19 pandemic progresses, research on mechanisms behind the transmission of SARS-CoV-2 shows conflicting evidence \cite{who2020report, Chinazzi395, guan2020clinical}. While effects of mobility have been extensively discussed, less is known on other factors such as changing awareness in the population
\cite{higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020}
or the effects of temperature
\cite{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020}. A limiting factor in many studies is the lack of a causal approach to
assess the causal contributions of various factors
\cite{Greenland1999}. This can lead to distorted estimates of the
causal factors with observational data
\cite{Greenland1999, schipf_directed_2011, textor_robust_2017}.

@@

Another problem with assessing the effects of various determinants of
SARS-CoV-2 spread is the heterogeneity of the countries and regions
examined for example in the Johns Hopkins University (JHU) COVID-19
database \cite{jhucovid19db2020}. The comparison of time series of case
numbers from different countries and observational periods can be
strongly distorted by different factors like testing capacities and
regional variations.

Our objective is to provide estimates of the causal effects of the main
drivers of the pandemic with reduced bias. We conducted a scoping review
of the available studies regarding signaling pathways and determinants
of the spread of SARS-CoV-2 infections and the reported new COVID-19
cases. Then we integrated the current findings into a directed acyclic
graph for the progress of the pandemic at the regional level. Using the
resulting model and the do-calculus we found identifiable effects
without blocked causal paths whose effects can be analyzed with
observational data \cite{Pearl_2014}. We used regional time series data
of all 401 German districts from various publicly available
sources to analyze these questions on a regional level. Germany is a
good choice in this regard, because it has ample data on contributing
factors on the regional level and has had high testing and treatment
capacities from early on in the pandemic.

\section*{Causal Model}\label{causal-model}

We used a directed acyclic graph (DAG)
\cite{schipf_directed_2011, textor_robust_2017} as a tool to analyze
the causal relationships between several exposures and SARS-CoV-2
spread. To get an overview on published associations, a scoping review
was conducted from 20th to 22nd of May 2020 within Pubmed and Google
scholar. Restrictions were applied to English and German language and
the publication date in the last one year. The following search terms
were applied to abstracts and title in Pubmed (``COVID-19'' OR
``COVID19'' OR ``Corona'' OR ``Coronavirus'' OR ``SARS-CoV-2'') and
connected separately in each case with the exposure variables
(``mobility'', ``public awareness'', ``awareness'', ``google
trends'',``ambient temperature'', ``temperature''). For ``mobility'', we
analyzed \(n=8\) studies, \(N=103\) were scanned in Pubmed, together
with the first ten pages (100 results) in Google scholar
(``awareness''/``public awareness''/``google trends'' \(n=9\),
\(N=215\); ``temperature''/``ambient temperature'' \(n=16\), \(N=235\)).
We integrated these findings where possible into the construction of our
DAG, which can be seen in Fig~\ref{fig:dag-covid-19}.

%\begin{figure}
%\centering
%\includegraphics{figures/f_full_covid19_dag.eps}
%\caption{\label{fig:dag-covid-19}DAG of determinants of reported COVID-19
%cases on the district level. Unobserved variables are light gray,
%variables marked with an asterisk (*) are confounded by
%weekday/holiday.}
%\end{figure}

\begin{figure}[!h]
\caption{{\bf DAG of determinants of reported COVID-19 cases on the district level.}
Unobserved variables are light gray, variables marked with an asterisk (*) are confounded by weekday/holiday.}
\label{fig:dag-covid-19}
\end{figure}

A number of studies report a strong association of \textbf{mobility}
restrictions on the number of new COVID-19 cases: Restrictive measures
(e.g. ``stay-at-home'' orders, travel bans, or school closures) are
shown to possibly reduce the COVID-19 incidence
\cite{chang_modeling_2020, Chinazzi395, fowler_effect_2020, kraemer_effect_2020, lasry_timing_2020, linka_outbreak_2020, mazzoli_effects_2020, xiong_data-driven_2020}.
However, some studies point out the combination of various
non-pharmaceutical interventions (NPIs) is decisive to prevent new
infections \cite{juni_impact_2020, lai_effect_2020}.

Google Trends \cite{google_trends} data can be used as a tool to get
insights into public interest (\textbf{awareness}) in the coronavirus
disease. Several recent studies imply a connection of relative search
volumes (RSV) indices and reported new COVID-19 cases
\cite{ayyoubzadeh_predicting_2020, effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, lin_google_2020, mavragani_tracking_2020, walker_use_2020, yuan_trends_2020, zhou_effects_2020}.
Some search terms e.g. ``COVID-19'' or ``coronavirus'' predated newly
infected cases/total number of cases by roughly 7 to 14 days for
different countries
\cite{effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020}.
Additionally, we acknowledged that individual risk-aware behavior might
be a reaction to the current COVID-19 burden (measured as reported cases
at the day of exposure).

Mixed evidence is available regarding the effect of
\textbf{temperature}: On the one hand several papers report an
association between increase in temperature and decrease in newly
infected COVID-19 cases
\cite{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
On the other hand, also the opposite has been found
\cite{auler_evidence_2020, xie_association_2020}. Some studies found no
association at all
\cite{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020}.
It should be noted that few studies considered other confounding
variables than meteorological ones (especially age and population
density among others
\cite{briz-redon_spatio-temporal_2020, juni_impact_2020, wu_effects_2020}).
In addition, the transferability of results between different climate
zones is questionable. To avoid possible bias caused by weather
variables other than temperature, we included rain, wind, and humidity
in our model.

When investigating causal determinants of SARS-CoV-2 infections, a
number of confounders have to be considered. Well-known risk factors for
SARS-CoV-2 as well as for other infections are demographic factors such
as age, gender, socio-economic status (SES), population density, and
foreign citizenship/ethnicity
\cite{de2020risk, Dragano2020.06.17.20133918, jhucovid19db2020}. In
Germany along with other countries (i.e.~Brazil, USA, or the UK),
populist parties or politicians and their electorate tend to be more
sceptical about effects of containment measures than the other part of
the electorate \cite{dohle_wingen_schreiber_2020, engle_staying_2020}.
Therefore we considered both ``right-wing populist party votes'' and
``voter turnout'' as possible confounders. Public health interventions
were also taken into account (contact restrictions, school closures
etc.), as their implementation showed strong correlations with
controlling the spread of SARS-CoV-2
\cite{cowling2020impact, juni_impact_2020, lai_effect_2020}. To avoid
bias due to reporting delay of case numbers we had to include weekday
and German holidays. We included some unobserved variables in our DAG
(e.g. ``Herd immunity''), too. Please note that ``Exposure to
SARS-CoV-2'' is itself an unobserved variable: German case numbers are
reported with delay after date of exposure and symptom onset.
\emph{Exposure to the virus} should not be confused with the formal
\emph{exposure variables} of the DAG (mobility, awareness, temperature).

%Lorem ipsum dolor sit~\cite{bib1} amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam Eq~(\ref{eq:schemeP}) sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id.~\cite{bib2} Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

%\begin{eqnarray}
%\label{eq:schemeP}
%	\mathrm{P_Y} = \underbrace{H(Y_n) - H(Y_n|\mathbf{V}^{Y}_{n})}_{S_Y} + \underbrace{H(Y_n|\mathbf{V}^{Y}_{n})- H(Y_n|\mathbf{V}^{X,Y}_{n})}_{T_{X\rightarrow Y}},
%\end{eqnarray}

\section*{Materials and methods}
\subsection*{Data}\label{data}

We collected and aggregated data on reported COVID-19 cases, regional
socio-demographic factors, weather, and general mobility on district and
state level in Germany for the period of 15 February 2020 to 8 July
2020. Our observation period for the outcome consisted of all dates from
23 February 2020 to 8 July 2020 (\(T=137\)), since we used a lag of 8
days for all confounders. We did not exclude any states or districts
(\(K=401\)). We analyzed the daily reported number of new cases as
outcome (\(K\cdot T=54{\ }937\) observations). The set of possible
predictors was derived from our causal DAG (see Table~\ref{tab:model-vars} and Fig~\ref{fig:dag-covid-19}). Due to
modelling and data limitations, some of the predictors were unobserved
or were modelled as a construct consisting of several variables. For our
causal analysis, we computed adjustment sets in three different
scenarios for separate exposures within the DAG: i) mobility of
population, ii) awareness of COVID-19 (i.e.~Google searches for
``corona''), iii) weather (i.e.~temperature).

\subsubsection*{Variables}\label{variables}

We downloaded German daily case numbers on district level reported by
Robert Koch Institute (RKI, \cite{casenumbers_rki}) and aggregated them by date. The number of daily active cases for
day \(d\) was derived by subtracting the total number of reported cases
on day \(d\) and day \(d-14\) (14 days as a conservative estimate for
the infectious period, which corresponds here to the required quarantine
time in Germany).
